---
title: "A robust pipeline for statistical metagenomic data analysis"
subtitle: "MISO Master Project | Metagenomics"
author: "Ines BAKLI, Alden SNEATH, Paul LEMONNIER"
format: 
  revealjs:
    theme: dark
    slide-number: true
    smaller: false
    scrollable: true
---

```{r setup}
#| include: false
library(igraph)
library(ggplot2)
library(here)
library(phyloseq)
library(knitr)
library(viridis) # AJOUT : Il faut charger la librairie pour les couleurs

# AJOUT : On remet les options globales de couleurs pour que ggplot s'en souvienne
theme_set(theme_minimal(base_size = 13))

options(
  ggplot2.discrete.fill = function(...)
    scale_fill_viridis_d(option = "cividis", ...),
  
  ggplot2.discrete.colour = function(...)
    scale_color_viridis_d(option = "cividis", ...)
)

load(here::here("posts", "data_preparation.RData"))
load(here::here("posts", "alpha_diversity.RData"))
load(here::here("posts", "beta_diversity.RData"))
load(here::here("posts", "Network.RData"))
```

# Introduction

## Metagenomic Challenge

::: {.incremental}

- Goal: Building a reproducible pipeline ➔ reliable conclusions on microbial interaction from metagenomic data.

- Data : Bacteria population abundance from patient's gut microbiome with differing diseases 

- Problem from microbiome sequencing : 
  - High dimensionnality
  - Sparsity
  - Compositionality
  
:::

## Dataset Structure {.smaller}

All analyses were performed using the `phyloseq` framework, integrating three essential matrices into a single `metagenomics` object:

::: {.panel-tabset}

### 1. OTU Table 
Counts/proportions of each taxon across samples.

```{r}
#| echo: false
# 1. Extrait la matrice d'OTU
otu_mat <- as(otu_table(metagenomics), "matrix")

# ✨ 2. L'ASTUCE : Trier les bactéries par leur abondance totale (décroissante)
# On calcule la somme de chaque ligne, et on ordonne la matrice
top_bacteries <- order(rowSums(otu_mat), decreasing = TRUE)
otu_mat_triee <- otu_mat[top_bacteries, ]

# 3. On sélectionne le Top 5 des bactéries et les 5 premiers patients
otu_preview <- otu_mat_triee[1:5, 1:5]

# 4. On nettoie les noms de lignes très longs (comme vu précédemment)
rownames(otu_preview) <- gsub(".*\\|[a-z]__", "", rownames(otu_preview))

# 5. On affiche le tableau
knitr::kable(otu_preview, row.names = TRUE, digits = 2)
```

### 2. Sample Data 
Clinical context and patient variables used for group comparisons

```{r}
#| echo: false
# 1. Extrait les méta-données en data.frame
meta_df <- as.data.frame(as(sample_data(metagenomics), "data.frame"))

# 2. Filtre pour ne garder que le premier individu de chaque groupe de maladie
# (Vérifie bien si ta colonne s'appelle "disease" ou "Disease_Status")
sample_preview <- meta_df[!duplicated(meta_df$disease), ]

# 3. Sélectionne les colonnes clés pour l'affichage sur la slide
# (Ajuste les noms de colonnes "age", "gender", "country" selon ton dataset)
colonnes_a_afficher <- c("age", "gender", "disease", "country")
knitr::kable(sample_preview[, colonnes_a_afficher], row.names = TRUE)
```


### 3. Taxonomy Table 
Biological identification for each OTU.
```{r}
#| echo: false
tax_df <- as.data.frame(as(tax_table(metagenomics), "matrix"))
tax_preview <- tax_df[!duplicated(tax_df$Phylum), ]

# ✨ LA LIGNE MAGIQUE : Nettoyer les noms de lignes
# On supprime tout ce qui précède le dernier caractère "|" suivi d'une lettre et de "__"
rownames(tax_preview) <- gsub(".*\\|[a-z]__", "", rownames(tax_preview))

# 3. Affiche les 5 premiers Phyla uniques pour que ça rentre bien sur la slide
# (Ajuste les noms de colonnes si besoin : "kingdom", "phylum", etc.)
colonnes_taxo <- c("Kingdom", "Phylum", "Class", "Order", "Family")
knitr::kable(head(tax_preview[, colonnes_taxo], 5), row.names = TRUE)
```

:::

# Data preparation

## 1. Quality control{.smaller}

::: {.incremental}

- Missing values : avoid analysis bias
- Distribution of sequencing depth = nb of reads/sample
:::
. . .
```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
bp_seq_depth
```
Allows the identifaction of outliers + normalisation

## 2. Data Overview {.smaller}

### Sample distribution {.smaller}
::: {.panel-tabset}
#### Gender {.smaller}
```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
Sample_dist_gender
```
#### Disease - Male {.smaller}
```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
SampleM_dist_disease
```
#### Disease - Female {.smaller}
```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
SampleF_dist_disease
```
:::

## 2. Data Overview {.smaller}

### Taxonomic composition {.smaller}
```{r}
#| cache: false
Top10b_disease
```
➔ Roughly same **microbial compositions**, differing **proportions**, also depending on the **nb of samples**

# Diversity analysis

## 1. Alpha diversity

Alpha diversity measures the diversity within a single sample. It can reflect both:

::: {.incremental}
-   **Richness**: the number of taxa present
-   **Evenness**: the distribution of abundances among taxa
:::

. . .

Several metrics can be used, we chose 4: Observed, Shannon, Simpson and Inverted Simpson


## Overview of the Alpha diversity
```{r}
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
AlphaD_combined
```

## Alpha diversity in details {.smaller}
::: {.panel-tabset}
### Oberved {.smaller}
Unique OTUs present per diseases
```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
AlphaD_Observed
```

### Shannon {.smaller}
Mesure the richness and evenness of species in a community
```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
AlphaD_Shannon
```

### Simpson {.smaller}
Probability : two random individuals from a community belong to the same species
```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
AlphaD_Simpson
```

### Inverted Simpson {.smaller}
Reciprocal of the Simpson = emphasize the importance of rare species in a community
```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
AlphaD_InvSimpson
```
:::
All of those differences in diversity needs to be **statistically significant**

## Statistical analysis {.smaller}
::: {.panel-tabset}
### 1. Normality test {.smaller}
To assess the normality, we used the Shapiro test
```{r}
knitr::kable(shapiro_table, 
              digits = 4, # Arrondir les p-values à 4 décimales
              col.names = c("Metric", "statistic", "P-value"))
```
➔ p-value < 0.05 = reject the normality ➔  Non parametric test

### 2. Significance test
To assess significant differences in microbial diversity across disease groups, we performed a Kruskal-Wallis test

```{r}
knitr::kable(kw_table, 
              digits = 4, 
              col.names = c("Metric", "statistic", "P-value"))
```

### 3. Pairwise comparaisons
KW = difference somewhere among the groups ➔  post-hoc pairwise comparisons to see which groups differ

```{r}
# # 1. Trier le tableau par la 4ème colonne (la p-value) par ordre décroissant
# PW_Wilcox_sorted <- PW_Wilcox_table[order(PW_Wilcox_table[, 4], decreasing = TRUE), ]
# 
# # 2. Extraire les 2 premières lignes et les 2 dernières lignes
# PW_Wilcox_extremes <- rbind(head(PW_Wilcox_sorted, 2), tail(PW_Wilcox_sorted, 2))
# 
# # 3. Afficher le tableau filtré avec kable
# knitr::kable(PW_Wilcox_extremes, 
#              digits = 4, 
#              row.names = FALSE, 
#              col.names = c("Metric", "Group A", "Group B", "P-value (BH)"))
```


```{r}
#| include: false
#library("here")
# load(here("posts", "datapreparation.RData"))
# load(here("posts", "alpha_diversity.RData"))
# load(here("posts", "beta_diversity.RData"))
# load(here("posts", "Network.RData"))
# 
# 
# knitr::kable(PW_Wilcox_table, 
#              digits = 4, # Arrondir les p-values à 4 décimales
#              col.names = c("Metric", "Group A", "Group B", "P-value (BH)"))
# set.seed(42)
# 
# plot(Network_phylum, 
#      layout = layout_with_fr(Network_phylum), 
#      vertex.size = 8,
#      vertex.color = "lightgrey", 
#      vertex.frame.color = "white",
#      vertex.label.color = "black", 
#      vertex.label.cex = 0.7,
#      edge.color = E(Network_phylum)$color, # Récupère les couleurs sauvegardées
#      edge.width = 1.5,
#      main = "Global Consensus Multi-Disease Network - Phylum")
# 
# # 1. Identifier les Hubs (les nœuds les plus connectés pour la Famille)
# degres_famille <- degree(Network_family)
# # On prend le Top 10 (ajuste ce chiffre si tu veux en montrer plus ou moins)
# seuil_top10 <- sort(degres_famille, decreasing = TRUE)[10] 
# 
# # 2. Préparer les étiquettes et les tailles
# # Base : pas de nom ("") et taille correcte pour les gris (4)
# labels_hubs_fam <- rep("", vcount(Network_family))
# tailles_noeuds_fam <- rep(4, vcount(Network_family))
# 
# # On ajoute le nom et on grossit (taille 8) UNIQUEMENT pour les hubs
# labels_hubs_fam[degres_famille >= seuil_top10] <- V(Network_family)$name[degres_famille >= seuil_top10]
# tailles_noeuds_fam[degres_famille >= seuil_top10] <- 8
# 
# # 3. Calcul de la disposition (Kamada-Kawai aère très bien le niveau Famille)
# set.seed(42)
# layout_famille <- layout_with_kk(Network_family)
# 
# # 4. Le dessin final
# plot(Network_family, 
#      layout = layout_famille, 
#      vertex.size = tailles_noeuds_fam,         
#      vertex.color = ifelse(degres_famille >= seuil_top10, "gold", "lightgrey"), 
#      vertex.frame.color = "white",
#      vertex.frame.width = 0.5,
#      vertex.label = labels_hubs_fam,           
#      vertex.label.color = "black", 
#      vertex.label.cex = 0.8,               
#      vertex.label.font = 2,                
#      vertex.label.dist = 0.8,              
#      edge.color = E(Network_family)$color,  
#      edge.width = 0.8,                     
#      main = "Global Consensus Multi-Disease Network - Family (Top Hubs Highlighted)")
# 
# # 1. Identifier les Hubs (les nœuds les plus connectés)
# degres_genre <- degree(Network_genre)
# # On trouve la valeur de connexion pour être dans le "Top 15"
# seuil_top15 <- sort(degres_genre, decreasing = TRUE)[15] 
# 
# # 2. Préparer les étiquettes et les tailles
# # Par défaut, pas d'étiquette ("") et petite taille (2)
# labels_hubs <- rep("", vcount(Network_genre))
# tailles_noeuds <- rep(4, vcount(Network_genre))
# 
# # On ne garde le nom et on grossit (taille 6) QUE pour les hubs
# labels_hubs[degres_genre >= seuil_top15] <- V(Network_genre)$name[degres_genre >= seuil_top15]
# tailles_noeuds[degres_genre >= seuil_top15] <- 6
# 
# # 3. Calcul de la disposition (graphopt est souvent performant pour les très gros réseaux)
# set.seed(42)
# layout_genre <- layout_with_graphopt(Network_genre)
# 
# # 4. Le dessin final !
# plot(Network_genre, 
#      layout = layout_genre, 
#      vertex.size = tailles_noeuds,         # Taille dynamique (gros pour hubs, petit pour le reste)
#      vertex.color = ifelse(degres_genre >= seuil_top15, "gold", "lightgrey"), # Les hubs en doré !
#      vertex.frame.color = "white",
#      vertex.frame.width = 0.2,
#      vertex.label = labels_hubs,           # Seulement les 15 noms apparaissent
#      vertex.label.color = "black", 
#      vertex.label.cex = 0.8,               # Texte bien lisible
#      vertex.label.font = 2,                # Texte en gras pour faire ressortir les hubs
#      vertex.label.dist = 0.8,              # Décale le texte
#      edge.color = E(Network_genre)$color,  # Bien utiliser les couleurs du genre
#      edge.width = 0.3,                     # Liens extrêmement fins pour voir à travers
#      main = "Global Consensus Multi-Disease Network - Genus (Top Hubs Highlighted)")
```
