---
title: "A robust pipeline for statistical metagenomic data analysis"
subtitle: "MISO Master Project | Metagenomics"
author: "Ines BAKLI, Alden SNEATH, Paul LEMONNIER"
format: 
  revealjs:
    theme: dark
    slide-number: true
    smaller: false
    scrollable: true
    zoom: true
---

```{r setup}
#| include: false
library(here)
source(here("functions", "inputData.R"))
load(here::here("posts", "datapreparation.RData"))
load(here::here("posts", "alpha_diversity.RData"))
load(here::here("posts", "beta_diversity.RData"))
load(here::here("posts", "Network.RData"))
```

# Introduction

## Metagenomic Challenge {.smaller}

::: incremental
-   Goal: Building a reproducible pipeline ➔ reliable conclusions on microbial interaction from metagenomic data.

-   Data : Bacteria population abundance from patients' gut microbiome with differing diseases

-   Problems from microbiome sequencing :

    -   High dimensionnality
    -   Sparsity
    -   Compositionality
:::

## Dataset Structure {.smaller}

All analyses were performed using the `phyloseq` framework, integrating three essential matrices into a single `metagenomics` object:

::: panel-tabset
### 1. OTU Table

Counts/proportions of each taxon across samples.

```{r}
#| echo: false
otu_mat <- as(otu_table(metagenomics), "matrix")

top_bacteries <- order(rowSums(otu_mat), decreasing = TRUE)
otu_mat_triee <- otu_mat[top_bacteries, ]

otu_preview <- otu_mat_triee[1:5, 1:5]

rownames(otu_preview) <- gsub(".*\\|[a-z]__", "", rownames(otu_preview))

knitr::kable(otu_preview, row.names = TRUE, digits = 2)
```

### 2. Sample Data

Clinical context and patient variables used for group comparisons

```{r}
#| echo: false
meta_df <- as.data.frame(as(sample_data(metagenomics), "data.frame"))

sample_preview <- meta_df[!duplicated(meta_df$disease), ]

colonnes_a_afficher <- c("age", "gender", "disease", "country")
knitr::kable(sample_preview[, colonnes_a_afficher], row.names = TRUE)
```

### 3. Taxonomy Table

Biological identification for each OTU.

```{r}
#| echo: false
tax_df <- as.data.frame(as(tax_table(metagenomics), "matrix"))
tax_preview <- tax_df[!duplicated(tax_df$Phylum), ]

rownames(tax_preview) <- gsub(".*\\|[a-z]__", "", rownames(tax_preview))

colonnes_taxo <- c("Kingdom", "Phylum", "Class", "Order", "Family")
knitr::kable(head(tax_preview[, colonnes_taxo], 5), row.names = TRUE)
```
:::

# Data preparation

## 1. Quality control {.smaller}

::: incremental
-   Missing values : avoid analysis bias
-   Distribution of sequencing depth = nb of reads/sample
:::

. . .

```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
bp_seq_depth
```

Allows the identifaction of outliers + normalisation

## 2. Data Overview {.smaller}

### Sample distribution {.smaller}

::: panel-tabset
#### Gender {.smaller}

```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
Sample_dist_gender
```

#### Disease - Male {.smaller}

```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
SampleM_dist_disease
```

#### Disease - Female {.smaller}

```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
SampleF_dist_disease
```
#### Country {.smaller}

```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
Sample_dist_country
```


#### Country - diseases {.smaller}

```{r}
#| cache: false
#| fig-height: 4.5
#| fig-align: center
#| out-width: "80%"
Sample_dist_country_disease
```

:::

## 2. Data Overview {.smaller}

### Taxonomic composition {.smaller}

```{r}
#| cache: false
Top10b_disease
```

➔ Roughly same **microbial compositions**, differing **proportions**, also depending on the **nb of samples**

# Diversity analysis

## 1. Alpha diversity {.smaller}

Alpha diversity measures the diversity within a single sample. It can reflect both:

::: incremental
-   **Richness**: the number of taxa present
-   **Evenness**: the distribution of abundances among taxa
:::

. . .

Several metrics can be used, we chose 4: Observed, Shannon, Simpson and Inverted Simpson

## Overview of the Alpha diversity {.smaller}

::: panel-tabset
### Oberved {.smaller}

Unique OTUs present per diseases

```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
p_observed
```

### Shannon {.smaller}

Mesure the richness and evenness of species in a community

```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
p_shannon
```

### Simpson {.smaller}

Probability : two random individuals from a community belong to the different species

```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
p_simpson
```

### Inverted Simpson {.smaller}

Reciprocal of the Simpson = emphasize the importance of rare species in a community

```{r}
#| fig-height: 4
#| fig-align: center
#| out-width: "80%"
p_invsimpson
```
:::

All of those differences in diversity needs to be **statistically significant**

## Statistical analysis {.smaller}

::: panel-tabset
### 1. Normality test {.smaller}

Assess the normality = Shapiro test

```{r}
ordre_metriques <- c("Observed", "Shannon", "Simpson", "InvSimpson")

shapiro_table <- shapiro_table %>%
  mutate(Metric = factor(Metric, levels = ordre_metriques)) %>%
  arrange(Metric) %>% 
  select(Metric, method, statistic, p.value)
knitr::kable(shapiro_table, 
              digits = 4, 
              )
```

➔ p-value \< 0.05 = reject the normality ➔ Non parametric test

### 2. Significance test

Assess the significant differences in microbial diversity across disease groups = Kruskal-Wallis test

```{r}
ordre_metriques <- c("Observed", "Shannon", "Simpson", "InvSimpson")

kw_table <- kw_table %>%
  mutate(Metric = factor(Metric, levels = ordre_metriques)) %>%
  arrange(Metric) %>% 
  select(Metric, method, parameter, statistic, p.value)
knitr::kable(kw_table, 
              digits = 4, 
              )
```

➔ p-value \< 0.05 = difference somewhere among the groups

### 3. Pairwise comparaisons

KW = difference somewhere among the groups ➔ post-hoc pairwise comparisons to see which groups differ

```{r}

PW_Wilcox_sorted <- PW_Wilcox_table[order(PW_Wilcox_table[[3]], decreasing = TRUE), ]


knitr::kable(PW_Wilcox_sorted,
             digits = 4,
             row.names = FALSE,
             col.names = c("Group A", "Group B", "P-value (BH)",  "Metric"))
```

-   p-value \< 0.05 = significative difference in diversity beetween the 2 communities
-   p-value \> 0.05 = no significative difference
:::

## 2. Beta diversity {.smaller}

Beta diversity measures differences in microbial composition between samples.

We used 2 metrics :

::: incremental
-   **Bray-Curtis** : Considers both the microbial communitites presence/absence and relative abundance beetween samples

-   **Jaccard** : Considers only presence/absence.
:::

. . .

To visualise those dispersions, we used PCoA

## 2. Beta diversity {.smaller}

::: panel-tabset
### Bray-Curtis PCoA

```{r}
p_bray
```

### Jaccard PCoA

```{r}
p_jaccard
```
:::

## Statistical Analysis {.smaller}

::: panel-tabset
### 1. Permutation test

Needed to verify the homogeneity of multivariate dispersions

```{r}
disp_results <- data.frame(
  Metric = c("Bray-Curtis", "Jaccard"),
  F_value = c(disp_test_bray$tab$F[1], disp_test_jaccard$tab$F[1]),
  P_value = c(disp_test_bray$tab$`Pr(>F)`[1], disp_test_jaccard$tab$`Pr(>F)`[1])
)

knitr::kable(disp_results, 
             digits = 3, # Arrondit à 3 décimales pour un affichage propre
             col.names = c("Distance Metric", "F-value", "P-value (999 perms)"))
```

➔ p-value \< 0.05 = heterogeneity in the groups multivariate dispersion

### 2. Permanova {.smaller}

Asses whether the centroids of the groups (disease status) are significantly different

```{r}
perm_result <- data.frame(
  Metric = c("Bray-Curtis", "Jaccard"),
  
  R2_percent = c(
    round(permanova_bray[1, "R2"] * 100, 2), 
    round(permanova_jaccard[1, "R2"] * 100, 2)
  ),
  
  F_value = c(permanova_bray[1, "F"], permanova_jaccard[1, "F"]),
  P_value = c(permanova_bray[1, "Pr(>F)"], permanova_jaccard[1, "Pr(>F)"])
)

knitr::kable(perm_result, 
             digits = 3, 
             col.names = c("Distance Metric", "Variation Explained (R² %)", "F-value", "P-value"))
```

➔ p-value \< 0.05 = Significant differences in community structure

➔ R2 Bray-Curtis \> R2 Jaccard = abundance differences are more discriminative than just presence/absence

### 3. Pairwise comparaison

```{r}
# Bray-Curtis table
bray_tab <- data.frame(
  Comparison = names(pairwise_bray$pairwise$observed),
  Observed_Difference = pairwise_bray$pairwise$observed,
  P_value = pairwise_bray$pairwise$permuted,
  Distance = "Bray-Curtis"
)

# Jaccard table
jaccard_tab <- data.frame(
  Comparison = names(pairwise_jaccard$pairwise$observed),
  Observed_Difference = pairwise_jaccard$pairwise$observed,
  P_value = pairwise_jaccard$pairwise$permuted,
  Distance = "Jaccard"
)

# Combine and display with kable
library(knitr)
combined_table <- rbind(bray_tab, jaccard_tab)

kable(combined_table, 
      caption = "Pairwise comparisons of multivariate dispersions (permutest.betadisper)",
      col.names = c("Comparison", "Observed diff.", "p-value", "Distance"),
      row.names = FALSE,
      digits = 4,
      booktabs = TRUE)
```

➔ Pairwise comparisons revealed that the IBD group had significantly different multivariate dispersion compared to all other diseases (p \< 0.05 for all comparisons)
:::

## Metagenomic networks analysis {.smaller}

How microbial populations reorganize, compete, or cooperate depending on the host’s disease.

We used a consensus network approach :

::: incremental
-   **SpiecEasi (MB)**: Meinshausen-Bühlmann Neighborhood Selection = Node to node interactions

-   **SpiecEasi (Glasso)**: Graphical Lasso = Global estimation

-   **SparCC**: Sparse Correlations for Compositional Data = global correlations
:::

. . .

An interaction is validated if confirmed by 2 of 3 methods

We tested 3 taxonomic levels : *Phylum*, *Family* and *Genus*

## 1. Phylum Level {.smaller}

::: panel-tabset

### Network

```{r}
plot(Network_phylum,
     layout = layout_with_fr(Network_phylum),
     vertex.size = 8,
     vertex.color = "lightgrey",
     vertex.frame.color = "white",
     vertex.label.color = "black",
     vertex.label.cex = 0.7,
     edge.color = E(Network_phylum)$color, # Récupère les couleurs sauvegardées
     edge.width = 1.5,
     main = "Global Consensus Multi-Disease Network - Phylum")

legend_data <- unique(data.frame(
  Disease = E(Network_phylum)$Disease, 
  Color = E(Network_phylum)$color
))

legend("bottomright", 
       title = "Disease", 
       legend = legend_data$Disease, 
       col = legend_data$Color, 
       lwd = 3,       # Épaisseur du trait dans la légende
       bty = "n",     # "n" pour ne pas dessiner de boîte moche autour de la légende
       cex = 0.8)     # Taille du texte
```

### Metrics

```{r}
print_network_stats <- function(network_list, tax_level) {
  stats_df <- do.call(rbind, lapply(names(network_list), function(g) {
    net <- network_list[[g]]
    if(is.null(net)) return(NULL)
    
    data.frame(
      Tax_Level       = tax_level,
      Disease_Cohort  = g,
      Nodes_Count     = vcount(net),
      Edges_Count     = ecount(net),
      Graph_Density   = round(edge_density(net), 3),
      Average_Degree  = round(mean(degree(net)), 2)
    )
  }))
  return(stats_df)
}
#| echo: false
knitr::kable(print_network_stats(networks_phylum, "Phylum"))
```

-   Low edge count (interactions)
-   Very low density

➔ grouping of populations, lack of resolution

:::

## 2. Family Level {.smaller}

::: panel-tabset
### Network

```{r}
degres_famille <- degree(Network_family)
seuil_top10 <- sort(degres_famille, decreasing = TRUE)[10]

labels_hubs_fam <- rep("", vcount(Network_family))
tailles_noeuds_fam <- rep(4, vcount(Network_family))

labels_hubs_fam[degres_famille >= seuil_top10] <- V(Network_family)$name[degres_famille >= seuil_top10]
tailles_noeuds_fam[degres_famille >= seuil_top10] <- 8

set.seed(42)
layout_famille <- layout_with_kk(Network_family)

plot(Network_family,
     layout = layout_famille,
     vertex.size = tailles_noeuds_fam,
     vertex.color = ifelse(degres_famille >= seuil_top10, "gold", "lightgrey"),
     vertex.frame.color = "white",
     vertex.frame.width = 0.5,
     vertex.label = labels_hubs_fam,
     vertex.label.color = "black",
     vertex.label.cex = 0.5,
     vertex.label.font = 1,
     vertex.label.dist = 0.8,
     edge.color = E(Network_family)$color,
     edge.width = 0.8,
     main = "Global Consensus Multi-Disease Network - Family (Top Hubs Highlighted)")

legend_data <- unique(data.frame(
  Disease = E(Network_family)$Disease, 
  Color = E(Network_family)$color
))

legend("bottomright", 
       title = "Disease", 
       legend = legend_data$Disease, 
       col = legend_data$Color, 
       lwd = 3,       # Épaisseur du trait dans la légende
       bty = "n",     # "n" pour ne pas dessiner de boîte moche autour de la légende
       cex = 0.8)     # Taille du texte
```

### Metrics

```{r}
print_network_stats <- function(network_list, tax_level) {
  stats_df <- do.call(rbind, lapply(names(network_list), function(g) {
    net <- network_list[[g]]
    if(is.null(net)) return(NULL)
    
    data.frame(
      Tax_Level       = tax_level,
      Disease_Cohort  = g,
      Nodes_Count     = vcount(net),
      Edges_Count     = ecount(net),
      Graph_Density   = round(edge_density(net), 3),
      Average_Degree  = round(mean(degree(net)), 2)
    )
  }))
  return(stats_df)
}
#| echo: false
knitr::kable(print_network_stats(networks_family, "Family"))
```

-   Increase in nodes and edges
-   Optimal graph density

➔ Network architecture = specific to each diseases
:::

## 3. Genus Level {.smaller}

::: panel-tabset
### Network

```{r}
degres_genre <- degree(Network_genus)
seuil_top15 <- sort(degres_genre, decreasing = TRUE)[15]

labels_hubs <- rep("", vcount(Network_genus))
tailles_noeuds <- rep(4, vcount(Network_genus))

labels_hubs[degres_genre >= seuil_top15] <- V(Network_genus)$name[degres_genre >= seuil_top15]
tailles_noeuds[degres_genre >= seuil_top15] <- 6

set.seed(42)
layout_genre <- layout_with_graphopt(Network_genus)

plot(Network_genus,
     layout = layout_genre,
     vertex.size = tailles_noeuds,         # Taille dynamique (gros pour hubs, petit pour le reste)
     vertex.color = ifelse(degres_genre >= seuil_top15, "gold", "lightgrey"), # Les hubs en doré !
     vertex.frame.color = "white",
     vertex.frame.width = 0.2,
     vertex.label = labels_hubs,           # Seulement les 15 noms apparaissent
     vertex.label.color = "black",
     vertex.label.cex = 0.5,               # Texte bien lisible
     vertex.label.font = 1,                # Texte en gras pour faire ressortir les hubs
     vertex.label.dist = 0.8,              # Décale le texte
     edge.color = E(Network_genus)$color,  # Bien utiliser les couleurs du genre
     edge.width = 0.3,                     # Liens extrêmement fins pour voir à travers
     main = "Global Consensus Multi-Disease Network - Genus (Top Hubs Highlighted)")


legend_data <- unique(data.frame(
  Disease = E(Network_genus)$Disease, 
  Color = E(Network_genus)$color
))

legend("bottomright", 
       title = "Disease", 
       legend = legend_data$Disease, 
       col = legend_data$Color, 
       lwd = 3,       # Épaisseur du trait dans la légende
       bty = "n",     # "n" pour ne pas dessiner de boîte moche autour de la légende
       cex = 0.8)     # Taille du texte
```

### Metrics

```{r}
print_network_stats <- function(network_list, tax_level) {
  stats_df <- do.call(rbind, lapply(names(network_list), function(g) {
    net <- network_list[[g]]
    if(is.null(net)) return(NULL)
    
    data.frame(
      Tax_Level       = tax_level,
      Disease_Cohort  = g,
      Nodes_Count     = vcount(net),
      Edges_Count     = ecount(net),
      Graph_Density   = round(edge_density(net), 3),
      Average_Degree  = round(mean(degree(net)), 2)
    )
  }))
  return(stats_df)
}
#| echo: false
knitr::kable(print_network_stats(networks_genus, "Genus"))
```

-   Too much nodes and edges
-   Decreasing density

➔ Topological collapse = a lot of fragmentation
:::

## Conclusion: A Robust Bioinformatics Pipeline {.smaller}

::: incremental
-   **End-to-End Reproducibility:** Seamless integration of metadata, taxonomy, and OTU counts within the `phyloseq` framework.
-   **Statistical Rigor:** Addressed inherent metagenomic biases (sparsity, zero-inflation, compositionality) via Centered Log-Ratio (CLR) transformations and non-parametric testing.
-   **Consensus Network Inference:** Mitigated algorithmic artifacts and false positives by enforcing a strict majority vote across three distinct methodologies (*SPIEC-EASI MB, SPIEC-EASI Glasso, and SparCC*).
-   **Scalable Architecture:** Delivered a fully automated, version-controlled analytical environment using R, `igraph`, and Quarto.
:::

# THANK YOU FOR YOUR ATTENTION !
