---
title: "alpha_diversity"
format: html
editor: visual
---

{{< include preliminary.qmd >}}

## **Transforming data**

### Relative Abundance

Data is already normalized with sample sums \~ 100

This is used for Shannon, Simpson and Inverse Simpson as they are robust to sequencing depth if using **relative abundances.**

### Normalize sequencing depth - rarefaction

**Rarefaction** (random subsampling without replacement).

In microbial ecology and amplicon sequencing (e.g., 16S rRNA gene sequencing) where samples often have vastly different numbers of reads. Rarefying aims to **make all samples have the same total number of reads** by randomly drawing a subset of reads from each sample **without replacement**. The chosen target depth is usually the smallest library size among all samples, so that no sample is asked to provide more reads than it originally had. This is only used for observed alpha diversity as rarefaction ensures each sample contributes the same number of reads.

```{r}
#| label: normalized data  
#| include: true
# metagenomics_rel <- transform_sample_counts(metagenomics, function(x) x / sum(x))
# metagenomics_rel

metagenomics_rare <- rarefy_even_depth(metagenomics, sample.size = min(sample_sums(metagenomics)))
metagenomics_rare
```

## Calculate alpha diversity

```{r}
# alpha diversity for Shannon, Simpson, InvSimpson
alpha_div <- estimate_richness(metagenomics, measures = c("Shannon", "Simpson", "InvSimpson"))
p1 <- plot_richness(metagenomics, x = "disease", measures = c("Shannon", "Simpson", "InvSimpson"))
p1
ggsave("figures/Alpha/alpha_diversity_shannon_simpson_invSimpson.pdf", p1)

alpha_obs <- estimate_richness(metagenomics_rare, measures = c("Observed"))
p2 <- plot_richness(metagenomics_rare, measures = c("Observed"))
p2
ggsave("figures/Alpha/alpha_diversity_observed.pdf", p2)

alpha_df <- cbind(alpha_div, alpha_obs)
alpha_df$SampleID <- rownames(alpha_div)
alpha_df$Disease <- sample_data(metagenomics)$disease

```

## Normality Testing

```{r}
# Check Normality (Shapiro-Wilk) to decide test 
shapiro.test(alpha_df$Shannon)
shapiro.test(alpha_df$Simpson)
shapiro.test(alpha_df$InvSimpson)
shapiro.test(alpha_df$Observed)

```

If p-value \> alpha 0.5, data is normal, use ANOVA. All test rejects the null hypothesis, so we use non-parametric test (Kruskal-Wallis).

```{r}
kruskal.test(Shannon ~ Disease, data = alpha_df)
kruskal.test(Observed ~ Disease, data = alpha_df)
kruskal.test(Simpson ~ Disease, data = alpha_df)
kruskal.test(InvSimpson ~ Disease, data = alpha_df)
```

## Alpha Diversity Visualization

```{r}
p3 <- ggplot(alpha_df, aes(x = Disease, y = Shannon, fill = Disease)) +
  geom_boxplot() +
  theme_minimal() +
  ylab("Shannon Diversity") +
  xlab("Disease") +
  ggtitle("Alpha Diversity (Shannon)")
ggsave("figures/Alpha/alpha_boxplot_shannon.pdf", p3)
p3

p4 <- ggplot(alpha_df, aes(x = Disease, y = Simpson, fill = Disease)) +
  geom_boxplot() +
  theme_minimal() +
  ylab("Simpson Diversity") +
  xlab("Disease") +
  ggtitle("Alpha Diversity (Simpson)")
ggsave("figures/Alpha/alpha_boxplot_simpson.pdf", p4)
p4

p5 <- ggplot(alpha_df, aes(x = Disease, y = InvSimpson, fill = Disease)) +
  geom_boxplot() +
  theme_minimal() +
  ylab("InvSimpson Diversity") +
  xlab("Disease") +
  ggtitle("Alpha Diversity (InvSimpson)")
ggsave("figures/Alpha/alpha_boxplot_InvSimpson.pdf", p5)
p5

p6 <- ggplot(alpha_df, aes(x = Disease, y = Observed, fill = Disease)) +
  geom_boxplot() +
  theme_minimal() +
  ylab("Observed Diversity") +
  xlab("Disease") +
  ggtitle("Alpha Diversity (Observed)")
ggsave("figures/Alpha/alpha_boxplot_Observed.pdf", p6)
p6
```

## Post-Hoc

Since Kruskal-Wallis tells you there’s a difference **somewhere among the groups**, we may want to do **post-hoc pairwise comparisons** to see which groups differ. For non-parametric tests, we can use:

```{r}
library(FSA)
# Pairwise Wilcoxon test with p-value adjustment
pairwise.wilcox.test(alpha_df$Observed, alpha_df$Disease, p.adjust.method = "BH") 
pairwise.wilcox.test(alpha_df$Simpson, alpha_df$Disease, p.adjust.method = "BH")
pairwise.wilcox.test(alpha_df$InvSimpson, alpha_df$Disease, p.adjust.method = "BH")
pairwise.wilcox.test(alpha_df$Shannon, alpha_df$Disease, p.adjust.method = "BH")
# p.adjust.method = "BH" → controls for multiple comparisons (Benjamini-Hochberg FDR)
```
