---
  title: "Statistical Analysis of Metagenomic Data"
  format: html
  editor: visual
---

```{r}
#| label: load-packages
#| include: false

library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
```

## Abstract

Metagenomic data analysis is crucial for understanding microbial communities. This document outlines the statistical methods used to analyze metagenomic datasets, including diversity metrics, differential abundance testing, and multivariate analyses.

## Introduction

Metagenomics involves the study of genetic material recovered directly from environmental samples. Analyzing such data requires specialized statistical techniques to account for its complexity and high dimensionality. This document provides an overview of the statistical methods commonly employed in metagenomic data analysis.

## Data

```{r}
#| label: load-data
#| include: true
otu_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "OTU_table")
tax_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Taxonomy")
samples_df <- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Sample_data")

# Convert data frames to matrices and set row names. Phyloseq objects need to have row.names
otu_mat <- otu_mat %>%
  tibble::column_to_rownames("OTU_ID") 
tax_mat <- tax_mat %>% 
  tibble::column_to_rownames("Tax_ID")
samples_df <- samples_df %>% 
  tibble::column_to_rownames("Sample_ID")

#Transform into matrixes otu and tax tables (sample table can be left as data frame)
otu_mat <- as.matrix(otu_mat)
tax_mat <- as.matrix(tax_mat)

# Transform to phyloseq objects
OTU = otu_table(otu_mat, taxa_are_rows = TRUE)
TAX = tax_table(tax_mat)
samples = sample_data(samples_df)

carbom <- phyloseq(OTU, TAX, samples)
carbom

sample_names(carbom)
rank_names(carbom)
sample_variables(carbom)

total = median(sample_sums(carbom))
# standf = function(x, t=total) round(t * (x / sum(x)))
# carbom = transform_sample_counts(carbom, standf)
```

## Statistical summary

```{r}
#| label: stats-summary
#| include: true

summary(carbom)

```

```{r}
theme_set(theme_bw())
gp.ch = subset_taxa(carbom, Phylum == "Bacteroidetes")
?subset_samples
carbom_france = subset_samples(carbom, country == "france")
carbom_france <- prune_taxa(
  taxa_sums(carbom_france) > 0,
  carbom_france
)
plot_bar(gp.ch)
p = plot_bar(carbom, "disease", fill="Phylum", facet_grid=gender~country) + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")

ggsave(
  filename = "Phylum abundance between country, sex and disease.png",
  plot = p,
  width = 8,
  height = 6,
  dpi = 300
)
tax <- tax_table(carbom)
phy <- tax[, "Phylum"]
mat_bin <- (otu_mat > 0) * 1
prev <- rowSums(mat_bin) / ncol(mat_bin)
df_prev <- data.frame(
  Phylum = phy,
  Prevalence = prev
)
df_prev_phy <- df_prev %>%
  group_by(Phylum) %>%
  summarise(Prevalence = mean(Prevalence)) %>%
  arrange(desc(Prevalence))
h <- ggplot(df_prev_phy, aes(x = reorder(Phylum, Prevalence), y = Prevalence)) + xlab("Phylum") +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal()
ggsave(
  filename = "Phylum prevalence.png",
  plot = h,
  width = 8,
  height = 6,
  dpi = 300
)
```
