---
  title: "Statistical Analysis of Metagenomic Data"
  format: html
  editor: visual
---

```{r}
#| label: load-packages
#| include: false

library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
```

## Abstract

Metagenomic data analysis is crucial for understanding microbial communities. This document outlines the statistical methods used to analyze metagenomic datasets, including diversity metrics, differential abundance testing, and multivariate analyses.

## Introduction

Metagenomics involves the study of genetic material recovered directly from environmental samples. Analyzing such data requires specialized statistical techniques to account for its complexity and high dimensionality. This document provides an overview of the statistical methods commonly employed in metagenomic data analysis.

## Data

```{r}
#| label: load-data
#| include: true
otu_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "OTU_table")
tax_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Taxonomy")
samples_df <- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Sample_data")

# Convert data frames to matrices and set row names. Phyloseq objects need to have row.names
otu_mat <- otu_mat %>%
  tibble::column_to_rownames("OTU_ID") 
tax_mat <- tax_mat %>% 
  tibble::column_to_rownames("Tax_ID")
samples_df <- samples_df %>% 
  tibble::column_to_rownames("Sample_ID")

#Transform into matrixes otu and tax tables (sample table can be left as data frame)
otu_mat <- as.matrix(otu_mat)
tax_mat <- as.matrix(tax_mat)

# Transform to phyloseq objects
OTU = otu_table(otu_mat, taxa_are_rows = TRUE)
TAX = tax_table(tax_mat)
samples = sample_data(samples_df)

carbom <- phyloseq(OTU, TAX, samples)
carbom

sample_names(carbom)
rank_names(carbom)
sample_variables(carbom)

total = median(sample_sums(carbom))
# standf = function(x, t=total) round(t * (x / sum(x)))
# carbom = transform_sample_counts(carbom, standf)
```

## Statistical summary

```{r}
#| label: stats-summary
#| include: true

summary(carbom)

```

```{r}
typeof(otu_mat)
#type is double (double precision floating point numbers)
```

```{r}
which(colSums(otu_mat != 0) == 0)
summary(colSums(otu_mat))
#there is no column of samples with only zeroes.

```

## Taxonomic composition

```{r}
carbom_phylum <- tax_glom(carbom, taxrank = "Phylum")
carbom_phylum_per <- transform_sample_counts(carbom_phylum, function(x) x / sum(x))
plot_bar(carbom_phylum_per, fill = "Phylum") +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Taxonomic composition at Phylum level") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
plot_bar(carbom_phylum_per, x = "disease", fill = "Phylum") +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(title = "Phylum composition by disease group")
```

```{r}
carbom_genus <- tax_glom(carbom, taxrank = "Genus")

carbom_genus_per <- transform_sample_counts(carbom_genus, function(x) x / sum(x))

# keep top 10 genera to avoid clutter
top_genera <- names(sort(taxa_sums(carbom_genus_per), decreasing = TRUE))[1:10]
carbom_genus_top <- prune_taxa(top_genera, carbom_genus_per)

plot_bar(carbom_genus_top, x = "disease", fill = "Genus") +
  geom_bar(stat = "identity", position = "stack") +
  theme_minimal() +
  labs(title = "Top 10 Genera Composition by Disease") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Beta Diversity - PCoA by gender

```{r}

pcoa <- ordinate(carbom, method = "PCoA", distance = "bray")

plot_ordination(carbom, pcoa, color = "gender") +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCoA - Bray-Curtis by gender")
```

## Beta Diversity - PCoA by gender and disease

```{r}
pcoagd <- ordinate(carbom, method = "PCoA", distance = "bray")

plot_ordination(carbom, pcoagd, color = "gender", shape = "disease") +
  geom_point(size = 3) +
  theme_minimal()
```

## Number of samples of diseases by gender

Females:

```{r}
meta_df <- data.frame(sample_data(carbom))

disease_counts <- meta_df %>%
  group_by(gender, disease) %>%
  summarise(Count = n())

ggplot(disease_counts %>% filter(gender == "female"),
       aes(x = disease, y = Count, fill = disease)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Number of Female Samples per Disease") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Males:

```{r}

ggplot(disease_counts %>% filter(gender == "male"),
       aes(x = disease, y = Count, fill = disease)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Number of male Samples per Disease") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
