---
  title: "Statistical Analysis of Metagenomic Data"
  format: html
  editor: visual
---

```{r}
#| label: load-packages
#| include: false

library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
```

## Abstract

Metagenomic data analysis is crucial for understanding microbial communities. This document outlines the statistical methods used to analyze metagenomic datasets, including diversity metrics, differential abundance testing, and multivariate analyses.

## Introduction

Metagenomics involves the study of genetic material recovered directly from environmental samples. Analyzing such data requires specialized statistical techniques to account for its complexity and high dimensionality. This document provides an overview of the statistical methods commonly employed in metagenomic data analysis.

## Data

```{r}
#| label: load-data
#| include: true
otu_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "OTU_table")
tax_mat<- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Taxonomy")
samples_df <- read_excel("../data/dataset_MISO_project.xlsx", sheet = "Sample_data")

# Convert data frames to matrices and set row names. Phyloseq objects need to have row.names
otu_mat <- otu_mat %>%
  tibble::column_to_rownames("OTU_ID") 
tax_mat <- tax_mat %>% 
  tibble::column_to_rownames("Tax_ID")
samples_df <- samples_df %>% 
  tibble::column_to_rownames("Sample_ID")

#Transform into matrixes otu and tax tables (sample table can be left as data frame)
otu_mat <- as.matrix(otu_mat)
tax_mat <- as.matrix(tax_mat)

# Transform to phyloseq objects
OTU = otu_table(otu_mat, taxa_are_rows = TRUE)
TAX = tax_table(tax_mat)
samples = sample_data(samples_df)

carbom <- phyloseq(OTU, TAX, samples)
carbom

sample_names(carbom)
rank_names(carbom)
sample_variables(carbom)

total = median(sample_sums(carbom))
# standf = function(x, t=total) round(t * (x / sum(x)))
# carbom = transform_sample_counts(carbom, standf)
```

## Statistical summary

```{r}
#| label: stats-summary
#| include: true

summary(carbom)

```

```{r}
theme_set(theme_bw())
gp.ch = subset_taxa(carbom, Phylum == "Bacteroidetes")
?subset_samples
carbom_france = subset_samples(carbom, country == "france")
carbom_france <- prune_taxa(
  taxa_sums(carbom_france) > 0,
  carbom_france
)
plot_bar(gp.ch)
p = plot_bar(carbom, "disease", fill="Phylum", facet_grid=gender~country) + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")

ggsave(
  filename = "Phylum abundance between country, sex and disease.png",
  plot = p,
  width = 8,
  height = 6,
  dpi = 300
)
tax <- tax_table(carbom)
phy <- tax[, "Phylum"]
mat_bin <- (otu_mat > 0) * 1
prev <- rowSums(mat_bin) / ncol(mat_bin)
df_prev <- data.frame(
  Phylum = phy,
  Prevalence = prev
)
df_prev_phy <- df_prev %>%
  group_by(Phylum) %>%
  summarise(Prevalence = mean(Prevalence)) %>%
  arrange(desc(Prevalence))
h <- ggplot(df_prev_phy, aes(x = reorder(Phylum, Prevalence), y = Prevalence)) + xlab("Phylum") +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal()
ggsave(
  filename = "Phylum prevalence.png",
  bg = "white",
  plot = h,
  width = 8,
  height = 6,
  dpi = 300
)
```

```{r}
# Statistiques préliminaires
taxa <- taxa_names(carbom)
head(taxa)
tax_df <- as.data.frame(tax_table(carbom))
colnames(tax_df)
ncol(tax_df)
colSums(is.na(tax_df))
colSums(tax_df == "")
tax_df[!complete.cases(tax_df), ]
str(tax_table(carbom))
#Ici on voit qu'on a aps de valeurs de taxonomie manquante, que tout les OTUs ont bien les 7 niveaux remplis

#Analyse statistqiue descriptive au nieavu des échantillons 
otu <- as.data.frame(otu_table(carbom))

# si les taxa ne sont pas en lignes
if (!taxa_are_rows(carbom)) {
  otu <- t(otu)
}
#somme totale des reads
library_size <- colSums(otu)
#Nombre d'OTU observés = richesse brute
richesse <- colSums(otu > 0)
#Abondance moyenne par échantillons
mean_abundance <- colMeans(otu)
#Ecart-type par échantillons
sd_abundance <- apply(otu, 2, sd)
#Tableau récapitulatif
stats_samples <- data.frame(
  LibrarySize = library_size,
  Richness = richesse,
  MeanAbundance = mean_abundance,
  SDAbundance = sd_abundance
)
```

```{r}
# 1️⃣ Extraire la matrice OTU
otu <- as.data.frame(otu_table(carbom))
if (!taxa_are_rows(carbom)) {
  otu <- t(otu)
}

# 2️⃣ Calculer stats par échantillon
library_size <- colSums(otu)
richness <- colSums(otu > 0)
mean_abundance <- colMeans(otu)
sd_abundance <- apply(otu, 2, sd)

stats_samples <- data.frame(
  LibrarySize = library_size,
  Richness = richness,
  MeanAbundance = mean_abundance,
  SDAbundance = sd_abundance
)

# Assigner les noms d'échantillons
rownames(stats_samples) <- colnames(otu)

# 3️⃣ Fonction pour détecter les outliers IQR
detect_iqr_outliers <- function(x) {
  Q1 <- quantile(x, 0.25)
  Q3 <- quantile(x, 0.75)
  IQR_val <- IQR(x)
  lower <- Q1 - 1.5 * IQR_val
  upper <- Q3 + 1.5 * IQR_val
  
  outlier_flag <- (x < lower) | (x > upper)
  outlier_samples <- names(x)[outlier_flag]
  
  list(
    outlier_flag = outlier_flag,
    outlier_samples = outlier_samples,
    lower = lower,
    upper = upper
  )
}

# 4️⃣ Appliquer pour chaque métrique et visualiser
metrics <- colnames(stats_samples)
par(mfrow = c(1,1))  # un seul plot  # 2x2 plots

for (m in metrics) {
  x <- stats_samples[[m]]
  names(x) <- rownames(stats_samples)
  res <- detect_iqr_outliers(x)
  
  # Marquer les outliers dans le tableau
  stats_samples[[paste0("Outlier_", m)]] <- res$outlier_flag
  
  # Boxplot avec bornes et points outliers
  boxplot(x, main=paste0(m, " distribution"),
          ylab=m, outline=FALSE)
  abline(h=res$lower, col="red", lty=2)
  abline(h=res$upper, col="red", lty=2)
  
  # Ajouter les points outliers et noms
  if(length(res$outlier_samples) > 0){
    points(which(res$outlier_flag), x[res$outlier_flag],
           col="blue", pch=19)
    text(which(res$outlier_flag), x[res$outlier_flag],
         labels=res$outlier_samples, pos=4, cex=0.7)
    cat("Outliers for", m, ":\n")
    print(res$outlier_samples)
  } else {
    cat("No outliers detected for", m, "\n")
  }
}

par(mfrow = c(1,1))  # remettre l’affichage normal

# 5️⃣ Tableau final
stats_samples
```

```{r}
par(mfrow=c(1,1))
boxplot(stats_samples$LibrarySize, main="Library Size")
boxplot(stats_samples$Richness, main="Observed OTU Richness")
boxplot(stats_samples$MeanAbundance, main="Mean OTU Abundance")
boxplot(stats_samples$SDAbundance, main="SD OTU Abundance")
par(mfrow=c(1,1))

# ordination PCoA Bray-Curtis
dist_bc <- distance(carbom, method="bray")
ordination <- ordinate(carbom, method="PCoA", distance=dist_bc)

p <- plot_ordination(carbom, ordination, color="disease")

# affichage à l'écran (optionnel)
p

# sauvegarde directe
chemin <- "/Users/paullemonnier/Desktop/Master_MISO/Master 1/S2/Projet BiLille/m1-project/graphs/PCoA_BrayCurtis.png"
# sauvegarder le graphique ggplot
ggsave(filename = chemin, plot = p, width = 8, height = 6, dpi = 300)


```
