---
title: "report"
format: html
editor: visual
---

# Abstract
Metagenomic data analysis is crucial for understanding microbial communities. This document outlines the statistical methods used to analyze metagenomic data sets, including diversity metrics, differential abundance testing, and multivariate analyses on the gene content of the human gut microbiome and the enzymes encoded by these genes. The data set retrieved from Kaggle uses the shotgun sequencing method to sequence the genomes of untargeted cells in a community in order to elucidate community composition and function. Since the activities of the vast majority of enzymes found in microbiomes are unknown, the use of shotgun metagenomics discovered how common antibiotic genes are in our gut bacteria.

# Dataset
<https://www.kaggle.com/datasets/antaresnyc/human-metagenomics/data>

```{r}
#| label: load packages
#| include: false
# Install required packages
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")

# Core packages
# BiocManager::install(c("phyloseq", "microbiome", "DESeq2"))
# install.packages(c("tidyverse", "vegan", "ggplot2", "patchwork", "RColorBrewer"))
# install.packages("metacoder")


# Load libraries
library(phyloseq)
library(microbiome)
library(tidyverse)
library(vegan)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library("readxl")
library(vegan)
```

```{r}

#| label: import-data
#| include: true

getwd()
# Import data from Excel
otu_mat <- read_excel("data/dataset_MISO_project.xlsx", sheet = "OTU_table")
tax_mat <- read_excel("data/dataset_MISO_project.xlsx", sheet = "Taxonomy")
sample_metadata <- read_excel("data/dataset_MISO_project.xlsx", sheet = "Sample_data")

# Convert to matrices/dataframes for phyloseq
# Assuming first column is OTU IDs
otu_mat <- as.data.frame(otu_mat)
rownames(otu_mat) <- otu_mat[,1]
otu_mat <- otu_mat[,-1]
otu_matrix <- as.matrix(otu_mat)

# Taxonomy table
tax_mat <- as.data.frame(tax_mat)
rownames(tax_mat) <- tax_mat[,1]
tax_mat <- tax_mat[,-1]
tax_matrix <- as.matrix(tax_mat)

# Sample data
sample_metadata <- as.data.frame(sample_metadata)
rownames(sample_metadata) <- sample_metadata[,1]
sample_metadata <- sample_metadata[,-1]

# Create phyloseq object
OTU <- otu_table(otu_matrix, taxa_are_rows = TRUE)
TAX <- tax_table(tax_matrix)
SAMPLES <- sample_data(sample_metadata)

# Combine into phyloseq object
physeq <- phyloseq(OTU, TAX, SAMPLES)

# Check the object
physeq
```

# Introduction

## What is Metagenomics?

Metagenomics is the study of genetic material recovered directly from environmental or clinical samples, bypassing the need to culture individual organisms in the laboratory . Instead of sequencing the genome of a single isolated species, metagenomics analyzes the collective genomes of entire microbial communities from samples such as soil, water, gut contents, or human skin

Analyzing such data requires specialized statistical techniques to account for its complexity and high dimensionality. This document provides an overview of the statistical methods commonly employed in metagenomic data analysis.

## What is 16S rRNA?

The 16S ribosomal RNA (16S rRNA) gene is a universal phylogenetic marker found in all bacteria and archaea. It's ideal for microbial community profiling because:

-   Universal presence: Found in all prokaryotes

-   Conserved regions: Allow Polymerase Chain Reaction (PCR) amplification across diverse species

-   Variable regions: Contain sufficient sequence variation to distinguish different taxa

-   Large public databases: Extensive reference libraries (SILVA, Greengenes, RDP) exist for taxonomic assignment

## Alpha Diversity

Alpha diversity measures the diversity within a single sample. Think of it as "within-sample" diversity.

Alpha diversity has two components:

-   Richness: The number of different species/OTUs/ASVs present

-   Evenness: How evenly the abundances are distributed among species

Several key indices are used to quantify alpha diversity: Shannon and Simpson

```{r}
#| label: alpha-diversity
#| fig-cap: "Alpha Diversity Metrics by Sample Group"


# First, check what variables are available in your sample data
cat("Available sample variables:\n")
print(sample_variables(physeq))

# Calculate alpha diversity (using only Shannon and Simpson data has decimals)
alpha_div <- plot_richness(physeq, 
                           measures = c("Shannon", "Simpson"),
                           color = "disease")  # Using disease as the grouping variable. Change to another variable if needed.

# Add boxplot overlay 
alpha_plot <- alpha_div + 
  geom_boxplot(aes(fill = disease), alpha = 0.5) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw(base_size = 12) +
  labs(title = "Alpha Diversity by disease Status",
       x = "disease Status",
       y = "Diversity Index Value") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

print(alpha_plot)


# Extract alpha diversity values for statistical testing
alpha_values <- estimate_richness(physeq, measures = c("Shannon", "Simpson"))
sample_df <- as(sample_data(physeq), "data.frame")
alpha_values$disease <- sample_df$disease

# ANOVA test 
anova_shannon <- aov(Shannon ~ disease, data = alpha_values)
anova_simpson <- aov(Simpson ~ disease, data = alpha_values)


# Kruskal-Wallis tests
shannon_test <- kruskal.test(Shannon ~ disease, data = alpha_values)
simpson_test <- kruskal.test(Simpson ~ disease, data = alpha_values)


# Print statistical results
cat("\n### Kruskal-Wallis results\n")

cat("**Shannon Diversity:** Kruskal-Wallis chi-squared = ", 
    round(shannon_test$statistic, 3), ", df = ", shannon_test$parameter,
    ", p-value = ", format(shannon_test$p.value, digits = 4), "\n\n", sep = "")

cat("**Simpson Diversity:** Kruskal-Wallis chi-squared = ", 
    round(simpson_test$statistic, 3), ", df = ", simpson_test$parameter,
    ", p-value = ", format(simpson_test$p.value, digits = 4), "\n", sep = "")

cat("\n### ANOVA Results\n")
summary(anova_shannon)
summary(anova_simpson)
```
Note: Since the shotgun metagenomic data was normalized (containing decimal values), alpha diversity was assessed using Shannon and Simpson indices, which are appropriate for relative abundance data. Richness-based metrics (Observed OTUs, Chao1) require integer count data and were therefore not included in this analysis. Statistical comparisons between disease groups were performed using the non-parametric Kruskal-Wallis test.

```{r}
#| label: alpha-diversity2
#| fig-cap: "Alpha Diversity Metrics by Sample Group"

library("vegan")

sample_df <- diversity(obj$data$otu_table[, sample_df$Sample_ID], 
                       index = "shannon")



```

### Interpretation
Both alpha diversity metrics show highly significant differences among the four disease groups:

- Shannon Diversity: Kruskal-Wallis test (p = 4.21e-14) and ANOVA (p = 2.69e-12) both confirm extremely significant differences

- Simpson Diversity: Kruskal-Wallis test (p = 1.93e-12) and ANOVA (p = 1.82e-06) also show highly significant differences

For Shannon Diversity:

- The disease grouping explains a substantial portion of the variance

- F-statistic = 20.12 indicates strong group separation

- Mean squares: Between groups (3.833) vs. Within groups (0.191) - groups differ by ~20x more than within-group variability, indicating a strong effect of disease status on Shannon diversity

For Simpson Diversity:

- It's more modest but still significant effect

- F-statistic = 10.1

- Mean squares: Between groups (0.062) vs. Within groups (0.006)

Both parametric (ANOVA) and non-parametric (Kruskal-Wallis) tests agree, strengthening confidence in the findings. This is important because Kruskal-Wallis doesn't assume normal distribution while ANOVA does.

Overall, the disease status significantly impacts gut microbiome diversity. Shannon diversity (which emphasizes richness and evenness) shows stronger group differences than Simpson diversity (which emphasizes dominance). This suggests the disease groups differ more in terms of overall species richness and evenness rather than just dominant species



## Beta Diversity

Beta diversity measures the dissimilarity between samples. Think of it as "between-sample" diversity.

Beta diversity answers questions like:

-   Are gut microbiomes more similar within the same individual than between different individuals?

-   Do treatment and control groups harbor distinct microbial communities?

Key Distance Metrics: Bray Curtis and Jaccard
```{r}
#| label: beta-diversity
#| fig-cap: "Beta Diversity Ordination Plots"
#| include: true

# Calculate distance matrices
dist_bray <- distance(physeq, method = "bray")
dist_jaccard <- distance(physeq, method = "jaccard")

# Perform ordination (PCoA/MDS)
ord_bray <- ordinate(physeq, method = "MDS", distance = dist_bray)
ord_jaccard <- ordinate(physeq, method = "MDS", distance = dist_jaccard)

# Create plots
p1 <- plot_ordination(physeq, ord_bray, color = "disease") +
  geom_point(size = 3) +
  stat_ellipse(aes(group = disease), linetype = 2) +
  theme_bw() +
  ggtitle("Bray-Curtis Dissimilarity") +
  theme(legend.position = "bottom")
print(p1)

p2 <- plot_ordination(physeq, ord_jaccard, color = "disease") +
  geom_point(size = 3) +
  stat_ellipse(aes(group = disease), linetype = 2) +
  theme_bw() +
  ggtitle("Jaccard Distance") +
  theme(legend.position = "bottom")
print(p2)


# PERMANOVA tests
# Bray-Curtis PERMANOVA
permanova_bray <- adonis2(dist_bray ~ disease, data = sample_df, permutations = 999)

# Jaccard PERMANOVA
permanova_jaccard <- adonis2(dist_jaccard ~ disease, data = sample_df, permutations = 999)

# Display results
cat("\n### PERMANOVA Results\n")
cat("**Bray-Curtis Dissimilarity:**\n")
cat("R² = ", round(permanova_bray$R2[1], 3), 
    ", F = ", round(permanova_bray$F[1], 2),
    ", p-value = ", format(permanova_bray$`Pr(>F)`[1], digits = 4), "\n\n", sep = "")

cat("**Jaccard Distance:**\n")
cat("R² = ", round(permanova_jaccard$R2[1], 3), 
    ", F = ", round(permanova_jaccard$F[1], 2),
    ", p-value = ", format(permanova_jaccard$`Pr(>F)`[1], digits = 4), "\n", sep = "")

# Beta dispersion test (assumption check)
disp_bray <- betadisper(dist_bray, sample_df$disease)
disp_test <- permutest(disp_bray)

cat("\n### Beta Dispersion Test (Homogeneity of Variances)\n")
cat("p-value = ", format(disp_test$tab$`Pr(>F)`[1], digits = 4), "\n")

```
Interpretation:
Bray-Curtis measures compositional dissimilarity based on both presence/absence and abundance, while Jaccard focuses solely on presence/absence. Permanova is used to test whether the centroids of the groups (disease status) are significantly different in the multivariate space defined by the distance metrics.

Both beta diversity metrics show significant compositional differences between disease groups:

* Bray-Curtis: R² = 0.087, F = 15.23, p = 0.001
  + Moderate effect size in microbiome studies.
  + Based on both presence/absence and abundance of species.
  + Disease status explains ~8.7% of the variation in microbial community structure.

* Jaccard: R² = 0.056, F = 9.41, p = 0.001
  + Smaller but still significant effect.
  + Based purely on presence/absence of species.
  + The lower R² suggests that abundance patterns (captured by Bray-Curtis) differentiate groups better than just species occurrence
